---
title: "HW3"
author: "Alexander Grunenwald, Logan Morgoff, Grant Gilly, Refat Zeyed"
date: "2024-10-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggthemes)
set.seed(42)
```


# Question 1
## Part a.)
$$
(Y\mid \frac{1}{\lambda}) \propto \lambda^{\frac{n}{2} + \alpha -1}\exp{-\frac{\lambda n v}{2} + \beta - 1} \mid v=\frac{1}{n}\sum(y_i-\mu)^2 \\
$$
$$
P(Y\mid\lambda^{-1}) \sim Gamma(\frac{n}{2} + \alpha, \frac{nv}{2} + \beta) \\
$$

## Part b.)

$$
\text{Given } y_i\in [8.4, 10.1, 9.4], \mu=8, \alpha=3, \beta=2, v=\frac{1}{3}\lvert\lvert y- \mu\lvert\lvert = 2.1766 \\
$$
$$
P(\lambda\mid y,\mu,\alpha,\beta) \sim Gamma(4.5, 5.265) \\
$$
$$
\mathbb{E}[\lambda\mid y] = \frac{4.5}{5.265} = 0.8547\\
$$

# Question 2


```{r}
crab <- read.table("data/Crabs.txt", header=T)
head(crab)
carapace <- crab$width
sample_mean <- mean(carapace)
sample_sd <- sqrt(var(carapace))
num_bins <- as.integer(sqrt(length(carapace)))

ggplot(data=crab, mapping = aes(x=carapace)) + geom_histogram(aes(y=after_stat(density)), bins = num_bins) +  stat_function(fun = dnorm, 
                args = list(mean = sample_mean, sd = sample_sd), 
                color = "red", linewidth = 1) + theme_economist()
  
```
The Normal distribution with the parameters of the smaple mean and variance seems to be a good fit to this distribution. 



Given that the prior and sampling distribution are Normal

$$
\newline(y\mid \mu\sigma^2)\sim N(\mu\mid y,\sigma^2=4.8)
$$
$$
(\mu\mid\mu_o=20,\tau_0^2=1.2)\sim N(20,1.2)
$$
Then the posterior distribution will be normal with the following parameters

$$
P(\mu\mid \sigma^2 = 4.8, y) \propto \exp[-\frac{1}{2}(\frac{\sum{(\mu-\mu_0)}}{\tau_0^2} + \frac{\sum({y_i -\mu})^2}{4.8})]
$$
$$
P(\mu\mid\bar{y}) = N(\mu \mid\mu_n,\tau_n^2) = N(26.1256, 0.02739267)
$$

```{r}
ggplot() + 
  stat_function(
    fun = dnorm,
    args = list(mean = 20, sd = 1.2),
    aes(color = "Prior"),
    linewidth = 1
  ) +
  stat_function(
    fun = dnorm,
    args = list(mean = 26.156, sd = 0.1655073),
    aes(color = "Posterior"),
    linewidth = 1
  ) +
  xlim(15, 30) +  
  ylim(0, 2.5) +
  labs(
    title = "Prior and Posterior Distributions",
    x = "x",
    y = "Density",
    color = "Legend"  
  ) +
  scale_color_manual(values = c("Prior" = "blue", "Posterior" = "red")) + 
  theme_economist()

26.156 + 0.1655073*(2)
26.156 - 0.1655073*(2)
```
I observe that the centrality of the posterior has shifted after the data update as well as the Margin of Error has significantly decreased in relation to the prior.  

# Question 3
### For School 1
```{r}
s1_school <- read.table("data/school1.txt")

# Param Decleration
mu_0 <- 5
kappa_0 <- 1
a <- 1
b <- 4

y1_bar <- mean(s1_school$V1)
n1 <- length(s1_school$V1)
s1_var <- var(s1_school$V1)

# Posterior Param Calculation
kappa_n <- kappa_0 + n1
mu_n <- (kappa_0 * mu_0 + n1 * y1_bar) / kappa_n
a_n <- a + n1 / 2
b_n <- b + 0.5 * sum((s1_school$V1 - y1_bar)^2)

# Sample from inverse Gamma
sigma2_s1_posterior <- 1 / rgamma(10000, shape = a_n, rate = b_n)

# Calculate Posterior Distribution for Means
sigma_s1_posterior <- sqrt(sigma2_s1_posterior)
mu_s1_posterior <- rnorm(10000, mean = mu_n, sd = sigma_s1_posterior / sqrt(kappa_n))

posterior_mean_s1 <- mean(mu_s1_posterior)
credible_interval_s1 <- quantile(mu_s1_posterior, probs = c(0.025, 0.975))

credible_interval_s1
posterior_mean_s1
```
### For School 2

```{r}
s2_school <- read.table("data/school2.txt")
# Param Decleration
mu_0 <- 5
kappa_0 <- 1
a <- 1
b <- 4

y2_bar <- mean(s2_school$V1)
n2 <- length(s2_school$V1)
s2_var <- var(s2_school$V1)

# Posterior Param Calculation
kappa_n <- kappa_0 + n2
mu_n <- (kappa_0 * mu_0 + n2 * y2_bar) / kappa_n
a_n <- a + n2 / 2
b_n <- b + 0.5 * sum((s2_school$V1 - y2_bar)^2)

# Sample from inverse Gamma
sigma2_s2_posterior <- 1 / rgamma(10000, shape = a_n, rate = b_n)

# Calculate Posterior Distribution for Means
sigma_s2_posterior <- sqrt(sigma2_s2_posterior)
mu_s2_posterior <- rnorm(10000, mean = mu_n, sd = sigma_s2_posterior / sqrt(kappa_n))

posterior_mean_s2 <- mean(mu_s2_posterior)
credible_interval_s2 <- quantile(mu_s2_posterior, probs = c(0.025, 0.975))

credible_interval_s2
posterior_mean_s2
```
### Sigma for School 1
```{r}
quantile(sigma_s1_posterior, probs = c(0.025, 0.975))
mean(sigma_s1_posterior)
```



### Sigma for School 2
```{r}
quantile(sigma_s2_posterior, probs = c(0.025, 0.975))
mean(sigma_s2_posterior)
```
### Probability that mu1 is greater than mu2
```{r}
mean(mu_s1_posterior > mu_s2_posterior)
```

